---
title: "README for Final assignment for the 'Getting and Cleaning data' Coursera class"
author: "louiis @ https://github.com/louiis/gettingandcleaningdata-wk4"
date: "February 16, 2016"
output: html_document
---

### Preliminary step: Loading libraries and data
``` {r}
library(dplyr)
library(plyr)
library(tidyr)
download.file(url = "https://d396qusza40orc.cloudfront.net/getdata%2Fprojectfiles%2FUCI%20HAR%20Dataset.zip",
              destfile = "data/wearabledata.zip",
              method = "curl")

unzip("data/wearabledata.zip", exdir = "./data/")
```

## Q1. Merge the test and train data frames
``` {r}
# get the text files to read
file_list_test <- list.files("./data/UCI HAR Dataset/test/", pattern = "*.txt")
file_list_train <- list.files("./data/UCI HAR Dataset/train/", pattern = "*.txt")

# inverse the order of the files, to put subject and activity columns upfront
file_list_test <- file_list_test[order(desc(file_list_test))]
file_list_train <- file_list_train[order(desc(file_list_train))]

# create test data frame
rm(test)
for (file in file_list_test){
    # if the merged dataset doesn't exist, create it
    if (!exists("test")){
        test <- read.table(paste0("./data/UCI HAR Dataset/test/",file))
    } else {
        temp_dataset <-read.table(paste0("./data/UCI HAR Dataset/test/",file))
        test<-cbind(test, temp_dataset)
        rm(temp_dataset)
    }
}

# create train data frame
rm(train)
for (file in file_list_train){
    # if the merged dataset doesn't exist, create it
    if (!exists("train")){
        train <- read.table(paste0("./data/UCI HAR Dataset/train/",file))
    } else {
        temp_dataset <-read.table(paste0("./data/UCI HAR Dataset/train/",file))
        train<-cbind(train, temp_dataset)
        rm(temp_dataset)
    }
}

# Merge the test and train data frames
merged <- rbind(test,train)

# change the column names
names(merged)[1] <- "activity"
names(merged)[2] <- "subject"

# Printout
str(merged)
```



## Q2. Extract the mean and standard deviation for each measurement
```{r}
# Read the names of the observations from data file
features <- read.table(file = "data/UCI HAR Dataset/features.txt", stringsAsFactors = FALSE)
featuretitles <- features[,2]

# Obtain which column number contain "mean" and "std"
tokeepmean <- which(sapply(featuretitles,grepl,pattern="mean"))
tokeepstd <- which(sapply(featuretitles,grepl,pattern="std"))

# Merge (concatenate) these column numbers, include 1 and 2 (activity and subject)
tokeep <- c(1,2,tokeepmean+2 ,tokeepstd+2)
tokeep <- tokeep[order(tokeep)]

# Select the approriate columns out of the "merged" data frame from Q1
df <- select(merged,tokeep)

# Printout
str(df)
```

## Q3. Replace activity values (1-6) by descriptive activity names in dataset
```{r}
# read activity labels from data file
activities <- read.table(file = "data/UCI HAR Dataset/activity_labels.txt")
activitylabel <- activities[,2]
head(activitylabel)

# Assign the activity label to corresponding activity value
df$activity <- sapply(df$activity,function(x) {activitylabel[x]})

# Printout 
str(df)
```

## Q4. Appropriately labels the data set with descriptive variable names.
This task was performed along Q1 and Q2
```{r}
# Printout
names(df)
```

## Q5. Create tidy dataset with average value for each variable for each activity and subject
```{r}
# Get the column names for the variables (observations)
variables <- names(df)[-c(1,2)]

# Perform aggregate with function "mean" on these variables, by columns "activity" and "subject"
av.df <- aggregate(df[variables],by = df[c("activity","subject")], FUN = mean, na.rm = TRUE)

# Rename the variable column names by adding "average"
names(av.df)[-c(1,2)] <- paste0("average[" , variables , "]")

# Final printout
str(av.df)
```

## Generating output data
```{r}
# Write in text file
write.table(av.df, row.names = FALSE,file = "step5-data.txt")
```
